---
title: "Translation of R to C++"
author: "Konrad Krämer"
date: "`r Sys.Date()`"
subtitle: "The R package ast2ast"

# https://quarto.org/docs/presentations/revealjs/themes.html
format: 
  revealjs:
    theme: moon
editor: visual
twitter: yes
---

## Motivation and aim

**Problem**: Using R functions which: 

::: {.incremental}

- <font size="5"> are called very often </font> 
- <font size="5"> are called by C or C++: </font> 
    * <font size="5"> expensive copying of memory from R to C/C++ and *vice versa* </font> 

:::

```{r, eval = TRUE, echo = FALSE}
library(Rcpp)
library(ast2ast)
library(r2sundials)
library(RcppXPtrUtils)
library(microbenchmark)

# R version
ti <- seq(0, 5, length.out=101)
p <- list(a = 2)
p <- c(nu = 2, a = 1)
y0 <- 0
frhs <- function(t, y, p, psens) {
  -p["nu"]*(y-p["a"])
} 

res_exp <- r2cvodes(y0, ti,
                    frhs, param = p)
attributes(res_exp) <- NULL

# External pointer
ptr_exp <- cppXPtr(code = '
int rhs_exp(double t, const vec &y,
            vec &ydot,
            RObject &param,
            NumericVector &psens) {
            
  double a = 1;
  double nu = 2;
  ydot[0] = -nu*(y[0] - a);
  return(CV_SUCCESS);
}
',
depends=c("RcppArmadillo",
          "r2sundials","rmumps"),
includes="using namespace arma;\n
          #include <r2sundials.h>",
cacheDir="lib", verbose=FALSE)

pv <- c(a = 1)
res_exp2 <- r2cvodes(y0, ti,
                     ptr_exp, param = pv)
attributes(res_exp2) <- NULL
```

```{Rcpp, eval = TRUE, echo = FALSE}
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::depends(rmumps)]]
// [[Rcpp::depends(r2sundials)]]
// [[Rcpp::depends(ast2ast)]]
// [[Rcpp::plugins("cpp17")]]
#include "etr.hpp"
#include "RcppArmadillo.h"
#include "r2sundials.h"
using namespace arma;
using namespace Rcpp;

typedef int (*fp)(double t, const vec &y,
             vec &ydot, RObject &param,
             NumericVector &psens);

typedef void (*user_fct)(sexp& y_,
              sexp& ydot_);
user_fct Fct;

int rhs_exp_wrapper(double t, const vec &y,
                    vec &ydot, RObject &param,
                    NumericVector &psens) {
  NumericVector p(param);
  const int size = y.size();
  sexp ydot_(size, ydot.memptr(), 2);
  
  double* ptr = const_cast<double*>(
                y.memptr());
  sexp y_(size, ptr, 2);
  Fct(y_, ydot_);
  return(CV_SUCCESS);
}


// [[Rcpp::export]]
NumericVector solve_ode(XPtr<user_fct> inp,
                        NumericVector time,
                        NumericVector y) {
Fct = *inp;
XPtr<fp> ptr = XPtr<fp>(new fp(
            &rhs_exp_wrapper));

Environment pkg = 
  Environment::namespace_env("r2sundials");
Function solve = pkg["r2cvodes"];
NumericVector output = solve(y, time,
                             ptr, time);

return output;
}
```

```{r, eval = TRUE, echo = FALSE}
out <- microbenchmark(
  r2cvodes(y0, ti,
          frhs, param = p),
  r2cvodes(y0, ti,
          ptr_exp, param = pv))

boxplot(out, names=c("R", "C++"))
```


## Motivation and aim {auto-animate=true}

::: {style="margin-top: 100px;"}
**How to solve the problem?**
:::

## Motivation and aim {auto-animate=true}

::: {style="margin-top: 100px;"}
**Solution**: Translating an R function using *ast2ast*
:::

::: {.incremental}

- Optimization
- ODE-functions

:::

```{r}
# ast2ast version
ti <- seq(0, 5, length.out=101)
y0 <- 0

library(ast2ast)
ode <- function(y, ydot) {
  nu <- 2
  a <- 1
  ydot[1] <- -nu*(y[1] - a)
}
pointer_to_ode <- translate(ode,
                          reference = TRUE)
res_exp3 <- solve_ode(pointer_to_ode,
                      ti, y0)
attributes(res_exp3) <- NULL

stopifnot(identical(res_exp,
                res_exp2,
                res_exp3))
out <- microbenchmark(
  r2cvodes(y0, ti,
          frhs, param = p),
  r2cvodes(y0, ti,
          ptr_exp, param = pv),
  solve_ode(pointer_to_ode,
          ti, y0))

boxplot(out, names=c("R", "C++", "ast2ast"))
```

## 1. Implemention of *ETR* {auto-animate="true"}

```{Rcpp, eval = TRUE, echo = FALSE}
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::depends(ast2ast)]]
#include "etr.hpp"
// [[Rcpp::plugins("cpp17")]]

// [[Rcpp::export]]
void example() {
  sexp pi = 3.14;
  sexp vec = colon(1, 4);
  sexp mat = matrix(5, 2, 2);
  print(pi*vec);
  print();
  print(vec + vec + mat);
}
```

- *Expression template library R (ETR)* 

:::: {.columns}

::: {.column width="50%"}
```{c, eval = FALSE, echo = TRUE}
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::depends(ast2ast)]]
#include "etr.hpp"
// [[Rcpp::plugins("cpp17")]]

// [[Rcpp::export]]
void example() {
  sexp pi = 3.14;
  sexp vec = colon(1, 4);
  sexp mat = matrix(5, 2, 2);
  print(pi*vec);
  print();
  print(vec + vec + mat);
}
```
:::

::: {.column width="50%"}
```{r, echo = TRUE}
example()
```
:::

::::

## 1. Comparison of *ETR* and R {auto-animate="true"}

:::: {.columns}

::: {.column width="50%"}
```{c, eval = FALSE, echo = TRUE}
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::depends(ast2ast)]]
#include "etr.hpp"
// [[Rcpp::plugins("cpp17")]]

// [[Rcpp::export]]
void example() {
  sexp pi = 3.14;
  sexp vec = colon(1, 4);
  sexp mat = matrix(5, 2, 2);
  print(pi*vec);
  print();
  print(vec + vec + mat);
}
```
:::

::: {.column width="50%"}
```{r, echo = TRUE, eval = TRUE}
example <- function() {
  pi = 3.14
  vec = 1:4
  mat = matrix(5, 2, 2)
  print(pi*vec)
  cat("\n")
  print(vec + vec + mat)
}
example()
```
:::

::::

::: {.incremental}

- C++ and R code are very similar
- R code is translated to C++

:::


## 2. Translation of R code {auto-animate="true"}

```{r, eval = FALSE, echo = TRUE}
a <- 1:4 - b
```


## 2. Translation of R code {auto-animate="true"}

:::: {.columns}

::: {.column width="50%"}
```{r, echo = TRUE}
lobstr::ast(a <- 1:4 - b)
```
:::

::: {.column width="50%"}
```{r, echo = TRUE}
lobstr::ast(a <- colon(1, 4) - b)
```
:::

::::




## Conclusion

**Thank you very much for your attention** \

**Get in contact:** \

- Github: <https://github.com/Konrad1991> 
- Twitter: <https://twitter.com/kraemer_konrad>

